---
title: "Resources"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
# code_folding: hide
---

<style type="text/css">
h1.title {
  font-size: 38px;
  font-family: "Tahoma", sans-serif;
  color: #1F3600; /* darkgreen from DietDiveR logo */
}
h1 { /* Header 1 */
  font-size: 28px;
  color: #1F3600;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: #1F3600;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: #1F3600;
}
body{ /* Normal  */
  background-color: rgb(251, 251, 251);
  }
blockquote{ 
  font-size: 14px;
}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #1F3600;
}
<!-- .navbar { -->
<!--   background-color: #882345; -->
<!--   padding-top: 0rem; -->
<!--   padding-bottom: 0rem; -->
<!-- } -->
</style>

<br>

Name the path to DietDiveR directory where input files are pulled.
```{r}
main_wd <- "~/GitHub/DietDiveR"
```

Load the necessary functions.
```{r}
source("lib/specify_data_dir.R")
source("lib/load_clean_ASA24.R")
source("lib/Food_tree_scripts/format.foods_2.r")
source("lib/QCOutliers.R")
source("lib/average.by.R")
```

You can come back to the main directory by:
```{r, eval=FALSE}
setwd(main_wd)
```

<br>

# Load ASA24 Items data

Specify the directory where the data is.
```{r, eval = FALSE}
SpecifyDataDirectory(directory.name= "eg_data/VVKAJ/")
```
<!-- [NOTE] SpecifyDataDirectory function is not executed by R markdown, but just showing  -->
<!--  the code so that this tutorial will have the same code as the R script.  -->

<!-- This is the R-markdown code to change directories. -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = 'eg_data/VVKAJ/')
```


Load the Items_f.txt file to take a look at it.  
You need the `quote=""` and `colClasses="character"` arguments to ignore quotation marks (do not regard them as a cell separator) and to load all the columns as characters so that FoodID will keep the trailing ".0".
```{r}
getwd()
ls()

```

Load your unprocessed (raw) food items-level data (as downloaded from the ASA24 study website).
The csv file will be loaded as a dataframe in R and be named as items_raw.
```{r}
items_raw <- read.csv("Raw_data/VVKAJ_Items.csv", sep = ",", header=T)
# items_raw <- read.table("VVKAJ_Items.txt", header = T)
```



# R and R-Studio

1. [R](http://www.r-project.org) 
2. [R-studio](http://www.rstudio.com)

# Other R resources

1. [Stackoverflow](https://stackoverflow.com) 
1. [The R Graph Gallery](https://r-graph-gallery.com) 
1. [RStudio Cheatsheets](https://support--rstudio-com.netlify.app/resources/cheatsheets/) 

<br>

# Dietary data analysis

## ASA24

1. [ASA24 Researchers Site](https://asa24.nci.nih.gov/researcherSite/ASA242013.aspx) 
1. [ASA24 Resources](https://epi.grants.cancer.gov/asa24/resources/) 
1. [Reviewing and clearning ASA24 data](https://epi.grants.cancer.gov/asa24/resources/cleaning.html)

## NHANES

1. [NHANES Top Page](https://www.cdc.gov/nchs/nhanes/index.htm) 
1. [NHANES Data & Documentation](https://wwwn.cdc.gov/nchs/nhanes/Default.aspx) 

<!-- This is to remove big white space at the end of the rendered html, which is produced due to toc_floating.  Source: https://stackoverflow.com/questions/52933437/how-to-remove-white-space-at-the-end-of-an-rmarkdown-html-output -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
