---
title: "Visualize ordination"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
# code_folding: hide
---

<style type="text/css">
h1.title {
  font-size: 38px;
  font-family: "Tahoma", sans-serif;
  color: #1F3600; /* darkgreen from DietDiveR logo */
}
h1 { /* Header 1 */
  font-size: 28px;
  color: #1F3600;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: #1F3600;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: #1F3600;
}
body{ /* Normal  */
  background-color: rgb(251, 251, 251);
  }
blockquote{ 
  font-size: 14px;
}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #1F3600;
}
</style>

<br>

# Introduction

There may be cases where you want to highlight specific users in the biplot because they have distinct metadata etc. Individuals can be highlighted with a thicker outline if desired. This can be done by loading the saved ordination results - Axis values and metadata combined, and the proportion of variance explained.

<br>

# Load functions and packages  

Name the path to DietDiveR directory where input files are pulled.
```{r}
main_wd <- "~/GitHub/DietDiveR"
```

Load necessary packages.
```{r}
library(ggplot2)
```

Load the necessary functions.
```{r}
source("lib/specify_data_dir.R")
source("lib/ggplot2themes.R")
```

Load the distinct 100 colors for use.
```{r}
distinct100colors <- readRDS("lib/distinct100colors.rda")
```

You can come back to the main directory by:
```{r, eval=FALSE}
setwd(main_wd)
```

Specify the directory where the data is.
```{r, eval=FALSE}
SpecifyDataDirectory(directory.name = "eg_data/NHANES")
```

```{r setup_1, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = 'eg_data/NHANES')
```

<br>

# Load ordination results

Change to the folder called "Ordination".
```{r, eval=FALSE}
SpecifyDataDirectory(directory.name = "eg_data/NHANES/Laboratory_data/Ordination/")

```

```{r setup_2, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = 'eg_data/NHANES/Laboratory_data/Ordination/')
```


Read in the metadata and users' Axis values.
```{r}
loaded_glu_w <- read.table("Food_D12_FC_QC_demo_QCed_males60to79_3Lv_ord_WEIGHTED_meta_users.txt", 
                           sep="\t", header=T)
```

Convert the GLU_index as a factor to plot it in order.
```{r}
loaded_glu_w$GLU_index <- factor(loaded_glu_w$GLU_index, levels= c("Normal", "Prediabetic", "Diabetic"))
```

Load the eigenvalues as a vector.
```{r}
eigen_loaded <- read.table("Food_D12_FC_QC_demo_QCed_males60to79_3Lv_ord_WEIGHTED_eigen.txt", header=T)
```

Make a vector that contains the variance explained.
```{r}
eigen_loaded_vec <- eigen_loaded[, 2]
```

Create a folder called "Viz_Ordination" to save the plots to be produced here.

<br>

---

# Highlight certain sample(s) - e.g. participants 83755 and 83820

Subset datapoint(s) that you would like to highlight.
```{r}
select_point <- subset(loaded_glu_w, SEQN=="83755" | SEQN=="83820" ) 
```

Plot participants in different colors, then plot the selected participants (SEQNs) above with a thicker outline.
```{r}
highlighted <- ggplot() +
  
  # Plot all the datapoints first.
  geom_point(loaded_glu_w, shape=21, size=3, alpha=1, colour="black",
             mapping=aes(x=Axis.1, y=Axis.2, fill=GLU_index)) +  
  scale_fill_manual(values = c("steelblue3", "yellow", "hotpink")) + 
  xlab( paste("Axis.1 (", paste(round(eigen_loaded_vec[1]*100, 1)), "%)", sep="") ) +
  ylab( paste("Axis.2 (", paste(round(eigen_loaded_vec[2]*100, 1)), "%)", sep="") ) +
  no_grid + space_axes + theme(aspect.ratio = 1) +
  
  # Add thicker outlined datapoints of selected individuals.
  geom_point(select_point, shape=21, size=3, alpha=1, stroke=2, color="black",
             mapping=aes(x=Axis.1, y=Axis.2, fill= GLU_index), show.legend=F) +
  
  # Add a caption to explain the highlighted datapoint(s).
  labs(caption="SEQN 83755 and 83820 are highlighted with a thick outline.") +
  theme(plot.caption =  element_text(hjust=1, face="italic"))
```

```{r,out.width="70%"}
highlighted
```

```{r}
ggsave("Viz_Ordination/Food_D12_FC_QC_demo_QCed_males60to79_3Lv_ord_WEIGHTED_Axis12_highlighted.pdf", 
       highlighted, device="pdf", width=7, height=5.5, unit="in", dpi=300)
```

<br>  

---

Come back to the main directory.
```{r, eval=FALSE}
setwd(main_wd)
```

<!-- This is to remove big white space at the end of the rendered html, which is produced due to toc_floating.  Source: https://stackoverflow.com/questions/52933437/how-to-remove-white-space-at-the-end-of-an-rmarkdown-html-output -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
