---
title: "Assign DivGroups by dietary diversity"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
# code_folding: hide
---

<style type="text/css">
h1.title {
  font-size: 38px;
  font-family: "Tahoma", sans-serif;
  color: #1F3600; /* darkgreen from DietDiveR logo */
}
h1 { /* Header 1 */
  font-size: 28px;
  color: #1F3600;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: #1F3600;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: #1F3600;
}
body{ /* Normal  */
  background-color: rgb(251, 251, 251);
  }
blockquote{ 
  font-size: 14px;
}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #1F3600;
}
</style>

<br>

# Introduction

Dietary diversity may have an important role in predicting health outcomes, and nuts/seeds/legumes food groups are likely to have health-promoting effects. To investigate the dietary diversity in specific food groups, we will select food items in a particular group, evaluate their diversity, and group participants based on their consumption patterns of the food group of interest.

<br>

# Load functions and packages  

Name the path to DietDiveR directory where input files are pulled.
```{r}
main_wd <- "~/GitHub/DietDiveR"
```

Load necessary packages.
```{r}
library(vegan)
library(reshape2)
library(data.tree)
```

Load necessary functions.
```{r}
source("lib/specify_data_dir.R")
source("lib/diversity_nth_tile.R")

# Load source scripts to build foodtrees and IFC tables.
source("lib/specify_data_dir.R")
source("lib/Food_tree_scripts/newick.tree.r")
source("lib/Food_tree_scripts/make.food.tree.r")
source("lib/Food_tree_scripts/make.food.ifc.r")
source("lib/Food_tree_scripts/make.fiber.ifc.r")
source("lib/Food_tree_scripts/make.dhydrt.ifc.r")
```

You can come back to the main directory by:
```{r, eval=FALSE}
setwd(main_wd)
```

Specify the directory where the data is.
```{r, eval=FALSE}
SpecifyDataDirectory(directory.name = "eg_data/NHANES/")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = 'eg_data/NHANES/')
```


<br>

# Select only food items with FoodCode starting with 4

Load averaged totals data, nutrition & food categories with demographic, gender-age, body measurements, and metadata. From NHANES_03_add_meta_GLU_index.R..

```{r}
totals <- read.delim("Total_D12_FC_QC_mean_QC_demo_ga_body_meta.txt")
```

This should have 4,164 people.
```{r}
nrow(totals)
```

<br>

---

Shannon's diversity is one of the measures to evaluate ecological diversity. We will use it to evaluate diet diversity of each individual in the NHANES data.

vegan::diversity calculates Shannon's diversity index one by one.

As an input, we need to have a table with SEQN as rows and food items as columns. So, we will create an IFC table first.

Load the food items data where QC-ed individuals were removed based on their totals data.
```{r}
food <- read.delim("Food_D12_FC_QC_demo_QCed.txt", sep= "\t", header=T)
```

Count the number of unique SEQNs. There should be 4164 people.
```{r}
length(unique(food$SEQN))
```

Here, we are interested in food items with their foodcode tarting from 4; nuts/seeds/legumes. Select only the rows that contain those food items.
```{r}
food4s <- subset(food, Food_code > 39999999 & Food_code < 50000000)
```

Check that the subsetted data only contains 4xxxxxxxs.
```{r}
summary(food4s$Food_code)
```

Check the summary of the subset data.
```{r}
paste(length(unique(food4s$SEQN)), "people consumed",
      nrow(food4s), "food items with duplicates.",
      "There are", length(unique(food4s$Food_code)), "unique food items.")
```

Save as a txt file.
```{r}
write.table(food4s, "Div/Food_D12_FC_QC_demo_QCed_4s.txt", sep= "\t", row.names=F,quote=F)
```

<br>

# Generate an IFC table

To calculate diversity of 4xxxxxxxs for each SEQN, we need to create an IFC table.

Specify where the data is.
```{r, eval=FALSE}
SpecifyDataDirectory("eg_data/NHANES/Div")
```

```{r setup 2, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = "eg_data/NHANES/Div")
```


```{r}
MakeFoodTree(nodes_fn="../../Food_tree_eg/NodeLabelsMCT.txt",
             addl_foods_fn = NULL,
             num_levels = 3,
             food_database_fn =            "Food_D12_FC_QC_demo_QCed_4s.txt",
             output_tree_fn =     "Foodtree/Food_D12_FC_QC_demo_QCed_4s_3Lv.nwk",
             output_taxonomy_fn = "Foodtree/Food_D12_FC_QC_demo_QCed_4s_3Lv.tax.txt"
)
```

<br>

---

Generate IFC tables for downstream analyses; IT MAY TAKE SOME TIME.

It is OK to see the following warning message:
  In write.table(fiber.ifc, output_fn, sep = "\t", quote = F, append = TRUE) :
  appending column names to file.

```{r}
MakeFoodIfc(food_records_fn=  "Food_D12_FC_QC_demo_QCed_4s.txt",
            food_record_id =  "SEQN",                              # The ID of your participants
            food_taxonomy_fn= "Foodtree/Food_D12_FC_QC_demo_QCed_4s_3Lv.tax.txt",       # Your taxonomy file produced by MakeFoodTree.
            output_fn =       "Foodtree/Food_D12_FC_QC_demo_QCed_4s_3Lv.food.ifc.txt")  # Output ifc file to be saved.
```

Load the generated IFC table.
```{r}
ifc <- read.delim("Foodtree/Food_D12_FC_QC_demo_QCed_4s_3Lv.food.ifc.txt")
```

It should have the dimension of number of unique foods x (1 food column + number of people + 1 taxonomy column).
243 x 2092, in this case.
```{r}
dim(ifc)
```

The column names have "X." at the beginning. We will take care of it later.
```{r}
ifc[1:4, 1:4]
```

<br>

# Calculate diversity of 4xxxxxxxs for each SEQN

Take out the foodID (description) and taxonomy from ifc.
```{r}
ifc2 <- ifc[, 2: (ncol(ifc)-1) ]
```

transpose so that the SEQN will come to rows.
```{r}
ifc2t <- as.data.frame(t(ifc2))
```

Add taxonomy as the column names of ifc2t.
```{r}
colnames(ifc2t) <- ifc$X.FOODID
```

Each row of ifc2t is SEQN. So, diversity needs to be calculated per each row.

Make a table to save results.
```{r}
SEQNdiv <- as.data.frame(matrix(nrow = nrow(ifc2t) , ncol = 4))
colnames(SEQNdiv) <- c("SEQN", "Shannon", "Simpson", "Invsimpson")
```

Do a loop to calculate Shannon's, Simpson, and inverse-Simpson diversity  for all SEQNs (in rows). This may take a few minutes.
```{r}
for( i in 1: nrow(ifc2t) ){
  SEQNdiv[i, 1] <- rownames(ifc2t)[i]
  SEQNdiv[i, 2] <- diversity(ifc2t[i, ], 'shannon')
  SEQNdiv[i, 3] <- diversity(ifc2t[i, ], 'simpson')
  SEQNdiv[i, 4] <- diversity(ifc2t[i, ], 'invsimpson')
}

head(SEQNdiv)
```

There should be no NA values.
```{r}
table(is.na(SEQNdiv), useNA="always")
```

Plot histograms of each of the diversity measures.
```{r}
par(mfrow = c(2, 2))
hist(SEQNdiv$Shannon, main="Shannon diversity", xlab="", breaks=10)
hist(SEQNdiv$Simpson, main="Simpson diversity", xlab="", breaks=10)
hist(SEQNdiv$Invsimpson, main="Invsimpson diversity", xlab="", breaks=10)
par(mfrow = c(1, 1))
```

Some have 0 diversity &rarr; If a row has only one non-zero values, then diversity will be zero. e.g. c(20,0,0,0,0,0,0).  i.e., those only consumed one nuts/seeds/legumes food have diversity of zero. None of them are normally distributed because of a lot of zero diversity values. For demonstration purposes, we will use Shannon’s diversity.

<br>

# Divide NHANES participants into four groups based on their diversity in nuts/seeds/legumes consumption

Our goal is to mark samples as:

DivGroup | Description | Shannon's diversity
---------- | -----------------| ------------------  
DivNA | Did not report any foods with Food ID of 4xxxxxxxx. | NA
Div0 | Reported 1 food with Food ID of 4xxxxxxx. | 0 
Div1 | Reported >1 foods with Food ID of 4xxxxxxx. lower. | > 0 
Div2 | Reported >1 foods with Food ID of 4xxxxxxx. upper. | > 0 
  
Remove the "X" in the SEQNdiv$SEQN for merging.
```{r}
SEQNdiv$SEQN <- gsub(SEQNdiv$SEQN, pattern = "X", replacement = "") 
```

First, need to add the diversity values to totals. Only take the rows present in both datasets.
```{r}
totals_div <- merge(totals, SEQNdiv, by='SEQN')
```

---

<br>

Select individuals whose diversity score is > 0, and group them into groups lower and upper (2-tiles) based on their Shannon's diversity measure.
```{r}
DivNthTile(input= totals_div, div.var="Shannon", nth.tile=2)
```

<br>

Define **Div1** and **Div2**. 
```{r}
out$DivGroup <- 
  ifelse(
    out$Div == 1,
    out$DivGroup <- 'Div1',
    out$DivGroup <- 'Div2'
  )
```

Select only the SEQN and DivGroup.
```{r}
SEQN_Div12 <- out[, c("SEQN", "DivGroup")]
```

<br>
  
Define **Div0**. 
```{r}
# Subset those that have Shannon index = 0.
totals_div_zero <- subset(totals_div, Shannon == 0)
```

Add DivGroup variable, and insert "Div0".
```{r}
totals_div_zero$DivGroup <- 'Div0'
```

Select only the SEQN and DivGroup.
```{r}
SEQN_Div0 <- totals_div_zero[, c("SEQN", "DivGroup")]
```

<br>

Define **DivNA**. 

Define "Not in" function.  By default it's not existent.
```{r}
`%!in%` <- Negate(`%in%`)
```

Subset those that are not in SEQNdiv.  
Those are the ones that did not consume nuts/seeds/legumes.
```{r}
totals_not_in_SEQNdiv <- totals[totals$SEQN %!in% SEQNdiv$SEQN, ]  
```

Add DivGroup variable, and insert "DivNA".
```{r}
totals_not_in_SEQNdiv$DivGroup <- 'DivNA'
```

Take only the SEQN and DivGroup.
```{r}
SEQN_DivNA <- totals_not_in_SEQNdiv[, c("SEQN", "DivGroup")]
```

---

<br>

Combine SEQN_DivNA, SEQN_Div0, and SEQN_Div12 for merging.
```{r}
SEQN_Div_NA_012 <- rbind(SEQN_DivNA, SEQN_Div0, SEQN_Div12)
```

Now, all the SEQNs have DivGroups.

Check that this should have the same number of rows as totals does. 
```{r}
identical(length(unique(SEQN_Div_NA_012$SEQN)), nrow(totals))
```

Merge DivGroups with the totals.
```{r}
totals_divgroup <- merge(totals, SEQN_Div_NA_012, all.x=T, by="SEQN")
```

Change DivGroup into a factor and specify the factor levels.
```{r}
totals_divgroup$DivGroup <- factor(totals_divgroup$DivGroup, 
                                   levels = c('DivNA', 'Div0', 'Div1', 'Div2') )
```

The individuals in totals were grouped into 4 groups depending on their consumption of 4xxxxxxx foods (or the lack thereof). The totals_divgroup has DivGroup column.
```{r}
table(totals_divgroup$DivGroup, useNA = "ifany")
```

Save the totals with DivGroup.
```{r}
write.table(totals_divgroup, "Total_D12_FC_QC_mean_QC_demo_ga_body_meta_DivGroup.txt",
            sep="\t", row.names=F, quote=F)
```

Health outcomes and demographic data could be explored by DivGroup.
  
<br>  

---

Come back to the main directory.
```{r, eval=FALSE}
setwd(main_wd)
```

<!-- This is to remove big white space at the end of the rendered html, which is produced due to toc_floating.  Source: https://stackoverflow.com/questions/52933437/how-to-remove-white-space-at-the-end-of-an-rmarkdown-html-output -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
