---
title: "Load and clean NHANES data"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 3
# code_folding: hide
---

<style type="text/css">
h1.title {
  font-size: 38px;
  font-family: "Tahoma", sans-serif;
  color: #1F3600; /* darkgreen from DietDiveR logo */
}
h1 { /* Header 1 */
  font-size: 28px;
  color: #1F3600;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: #1F3600;
}
h3 { /* Header 3 */
  font-size: 18px;
  color: #1F3600;
}
body{ /* Normal  */
  background-color: rgb(251, 251, 251);
  }
blockquote{ 
  font-size: 14px;
}
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #1F3600;
}
</style>

<br>

# Introduction

In this tutorial, we aim to demonstrate the utility of the DietDiveR package by applying it to the dietary data available from the National Health and Nutrition Examination Survey (NHANES).

NHANES is a national survey conducted with some thousands of participants across the US over two days. NHANES data is published every other year and consists of various kinds of data including dietary intake, body measurements, laboratory-taken data. There are numerous questions that can be asked with these data to connect dietary intake with various biomarkers. Here, as an example, we will explore the dietary intake (food items) of participants and look at the relationship between dietary patterns and diabetic status in NHANES 2015-2016.

Refer to the [NHANES data and documentation](https://wwwn.cdc.gov/nchs/nhanes/Default.aspx) for more information about each release (version) of NHANES and the variables measured.

Sample weight is an important tool to use if you aim for population-level estimates because the composition of demographics of NHANES participants differ slightly from that of the actual US population (i.e., certain demographic subgroups were oversampled). Weights are applied to adjust NHANES data to reflect the proportions of demographic subgroups in the actual population. However, this part is omitted in this tutorial as our main goal here is to demonstrate how to analyze NHANES data as is, not to make population-wise inferences.  For more detailed information, please refer to [the sample weight tutorial by NHANES](https://wwwn.cdc.gov/nchs/nhanes/tutorials/Weighting.aspx). 

Data from NHANES are in .XPT format, and you will need a specific R package, foreign, to import it into R. 
We will download the example data files from the [NHANES 2015-2016 page](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2015) and save them in the "Raw_data" folder in the "NHANES" folder.

Open "02_load_clean_NHANES_food_1.R" script (this script) in the "users" folder and load the packages needed as per the directions.

In this script, you will learn how to download datasets and associated other databases from NHANES, how to process that data with DietDiveR, and the food tree to generate dietary patterns, and how to integrate these dietary patterns with other biomarker data from NHANES to explore specific hypotheses.

<br>

# Load functions and packages  

Name the path to DietDiveR directory where input files are pulled.
```{r}
main_wd <- "~/GitHub/DietDiveR"
```

Install the foreign package if it is not installed yet.
```{r, eval=FALSE}
if (!require("foreign", quietly = TRUE)) install.packages("foreign")
```

<!-- Load foreign in the background-->
```{r, echo=FALSE}
library(foreign)
```

Load the necessary functions.
```{r}
source("lib/specify_data_dir.R")
source("lib/load_clean_NHANES.R")
source("lib/Food_tree_scripts/format.foods_2.r")
```

You can come back to the main directory by:
```{r, eval=FALSE}
setwd(main_wd)
```

Specify the directory where the data is.
```{r, eval=FALSE}
SpecifyDataDirectory(directory.name = "eg_data/NHANES")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = 'eg_data/NHANES')
```

<br>

# Prepare food codes, FPED, and food items files

## Load and prepare food codes 

Download food code data from the NHANES website and save it in "Raw_data" folder.  
Name the file and destination. mod="wb" is needed for Windows OS. Other OS users may need to delete it.
```{r, eval=FALSE}
download.file("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DRXFCD_I.XPT", 
              destfile = "Raw_data/FoodCodes_DRXFCD_I.XPT", mode="wb")
```

## Prepare the food code table - replace special characters with "_" or "and".
  
Format the food table and save it as a .txt file.
```{r}
PrepareFoodCodeTable(raw.food.code.table = "Raw_data/FoodCodes_DRXFCD_I.XPT",
                     out.fn = "FoodCodes_DRXFCD_I_f.txt")
```

Load the formatted foodcode table.
```{r}
foodcodetable_f <- read.table("FoodCodes_DRXFCD_I_f.txt", sep="\t", header=T)
```

Show the first 10 rows of the output and ensure special characters are gone; e.g., MILK, REDUCED FAT (2%) is now MILK, REDUCED FAT (2_).
```{r}
foodcodetable_f[1:10, ]
```

<br>

# Load and prepare Food Patterns Equivalent Database for use

The Food Patterns Equivalents Database (FPED) has the composition of nutrients and food categories of each food item coded by food codes. FPED can be downloaded from [USDA -ARS FPED databases](https://www.ars.usda.gov/northeast-area/beltsville-md-bhnrc/beltsville-human-nutrition-research-center/food-surveys-research-group/docs/fped-databases/).

You need to use the version of FPED that corresponds with the NHANES version that you explore. When using NHANES, you will need to use the FPED with the same release year as the year of NHANES you are analyzing. For this tutorial, FPED was downloaded from NHANES the FPED table is formatted by renaming the variables with R-loadable ones (e.g., "F_CITMLB (cup eq.)" &rarr; "F_CITMLB" and saved as "FPED_1516_forR.txt".

Load FPED15-16, needed for the AddFoodCat function.
```{r}
FPED <- read.table("FPED/FPED_1516_forR.txt", sep="\t", header=T)
```

Check the first 2 rows of FPED.
```{r}
head(FPED, 2)
```

Important! Change the food code column name as Food_code.
```{r}
colnames(FPED)[1] <- "Food_code"
```

<br>

## Load "food items" data and add food descriptions
Download food items data

Food data can be downloaded from the NHANES website. For each cycle of NHANES there are two days of dietary records available for many participants. The food-level data is available for each of these days of dietary records. These are stored separately as Day 1 and Day 2. Here we will download both days and combine them to obtain better estimates of diet.

Download day 1 data.
```{r, eval=FALSE}
download.file("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DR1IFF_I.XPT", 
              destfile = "Raw_data/DR1IFF_I.XPT", mode="wb")
```

Download day 2 data.
```{r, eval=FALSE}
download.file("https://wwwn.cdc.gov/Nchs/Nhanes/2015-2016/DR2IFF_I.XPT",
              destfile = "Raw_data/DR2IFF_I.XPT", mode="wb")
```


> **[NOTE]** Different alphabets are used on the variables' names in different release of NHANES data. Therefore, you will need to change the alphabet (and potentially the other parts of the variable names) in order to run this script with other releases of NHANES. For example, `DR1IFDCD` is a column name for the food code used in the NHANES 2015-2016, and the alphabet for this release is `I`. The alphabet for NHANES 2017-2018 is `J`, so the food code column in that dataset is DR1JFDCD, and the other columns in it have `DR1J` or `DR2J` as prefixes.  
>
> Whilst we will only use one dataset (2015-16) in this tutorial, you may want to combine several releases to analyze long-term trends. It is recommended that you check the variable names and food item names as the food databases are updated regularly, and they may not perfectly match across versions.

<br>

## Load and prepare Day 1 food items data
  
Import items data Day 1, add food item descriptions, and save it as a txt file. **OUTPUT WILL LIKELY BE A VERY LARGE FILE.**
```{r}
ImportNHANESFoodItems(data.name = "Raw_data/DR1IFF_I.XPT", 
                      food.code.column = "DR1IFDCD", 
                      food.code.table = foodcodetable_f,
                      out.fn = "DR1IFF_I_d.txt") # 'd' stands for food descriptions
```

Load the saved food items file.
```{r}
Food_D1 <- read.table("DR1IFF_I_d.txt", sep="\t", header=T)
```

Count the number of participants - should be 8,505 people.
```{r}
length(unique(Food_D1$SEQN)) 
```

Add the food category info and serving for each item. **WILL TAKE A FEW MOMENTS.**  
It is OK to see a message saying: `NAs introduced by coercion`. NAs will be removed later in the filtering process.
```{r}
AddFoodCat(input.food = Food_D1,
           fped = FPED,
           grams = "DR1IGRMS", 
           out.fn = "Food_D1_FC.txt")
```

<br>

## Load and prepare Day 2 food items data
  
Import items data Day 2, add food item descriptions, and save it as a txt file.
```{r}
ImportNHANESFoodItems(data.name = "Raw_data/DR2IFF_I.XPT",
                      food.code.column = "DR2IFDCD",
                      food.code.table = foodcodetable_f,
                      out.fn = "DR2IFF_I_d.txt")
```

Add food item description and save it as a txt file.
```{r}
Food_D2 <- read.table("DR2IFF_I_d.txt", sep="\t", header=T)
```

Count the number of participants - should be 7,027 people.
```{r}
length(unique(Food_D2$SEQN)) 
```

Do the same for Day 2. Add the food items info and serving for each item. **WILL TAKE A FEW MOMENTS.**  
It is OK to see a message saying: `NAs introduced by coercion`. NAs will be removed later in the filtering process.  
```{r}
AddFoodCat(input.food = Food_D2, 
             fped = FPED, 
             grams = "DR2IGRMS", 
             out.fn = "Food_D2_FC.txt")
```

<br>

# Load the Food_Dx_FC which has the food category data 

## Change column names to more intuitive ones
  
Food Day 1 with Food Category **WILL BE A VERY LARGE TABLE.**
```{r}
Food_D1_FC <- read.table("Food_D1_FC.txt", sep="\t", header=T)
```

Food Day 2 with Food Category **WILL BE A VERY LARGE TABLE.**
```{r}
Food_D2_FC <- read.table("Food_D2_FC.txt", sep="\t", header=T)
```

Change the colnames for downstream analyses.
```{r}
names(Food_D1_FC)[names(Food_D1_FC) == "DR1IFDCD"] <- "FoodCode"
names(Food_D1_FC)[names(Food_D1_FC) == "DR1IGRMS"] <- "FoodAmt"
names(Food_D1_FC)[names(Food_D1_FC) == "DRXFCLD"] <- "Main.food.description"
```

```{r}
names(Food_D2_FC)[names(Food_D2_FC) == "DR2IFDCD"] <- "FoodCode"
names(Food_D2_FC)[names(Food_D2_FC) == "DR2IGRMS"] <- "FoodAmt"
names(Food_D2_FC)[names(Food_D2_FC) == "DRXFCLD"] <- "Main.food.description"
```

Ensure the column names are changed.

Among the colnames of Food_D1_FC, show one that matches with "DR1IFDCD", if it exists.  
If it returns the specified colname, "DR1IFDCD", a column with that name exists.  
If it returns "character(0)", no colname matched the specified characters, which means the colname of our interest does not exist in the specified dataframe.
```{r}
names(Food_D1_FC)[names(Food_D1_FC) == "FoodCode"]
```

For day 2 dataset.
```{r}
names(Food_D2_FC)[names(Food_D2_FC) == "Main.food.description"]
```

Save after changing the column names. "cc" stands for column names changed.
```{r}
write.table(Food_D1_FC, "Food_D1_FC_cc.txt", sep="\t", row.names=F, quote=F)
```

```{r}
write.table(Food_D2_FC, "Food_D2_FC_cc.txt", sep="\t", row.names=F, quote=F)
```

<br>

## Replace special characters with "_" using FormatFoods.
  
`FotmatFoods()` function adds "Main.Food.Description" where special characters are removed/replaced, the previous Main.Food.Description as Old.Main.Food.Description, ModCode, and FoodID. $FoodID is a character vector, but has .0 at the end. 
```{r}
FormatFoods(input_fn="Food_D1_FC_cc.txt", output_fn= "Food_D1_FC_cc_f.txt", dedupe=F)
```

```{r}
FormatFoods(input_fn="Food_D2_FC_cc.txt", output_fn= "Food_D2_FC_cc_f.txt", dedupe=F)
```

<br>  

---

Come back to the main directory.
```{r, eval=FALSE}
setwd(main_wd)
```

<!-- This is to remove big white space at the end of the rendered html, which is produced due to toc_floating.  Source: https://stackoverflow.com/questions/52933437/how-to-remove-white-space-at-the-end-of-an-rmarkdown-html-output -->
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
